---
title: "paxcan_pca_pcoa_dapc_analyses"
author: "Isaac Buabeng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## load the necessary libraries
```{r}
library(adegenet)
library(pegas)
library(vcfR)
library(ape)
library(dartR)
library(plotly)
library(poppr) 
library(officer)
library(ggplot2)
library(tidyverse)
library(ggrepel)
# sourced the ggDAPC script from "https://raw.githubusercontent.com/wilsonfrantine/ggDAPC/main/ggDAPC.R" for DAPC visualizations
```



# load in and convert the data into a genlight object
```{r}
vcf <- read.vcfR("C:/Users/User/Documents/paxcan_project/r_based_analyses/14pops_biallelic_snps_miss0.4_thin10.newID.mac1.maf0.01.ldkb1r0.8.vcf")

# convert the vcf file into a genlight object
paxcan_gl <- vcfR2genlight(vcf)
```



# Filter loci and individuals
```{r}
# Ensure all required metadata and flags exist
paxcan_gl <- gl.compliance.check(paxcan_gl)   # adds missing flag skeletons

# Remove (or confirm absence of) monomorphic loci and set the flag
paxcan_gl <- gl.filter.monomorphs(paxcan_gl, verbose = 3)
```



# load in population data
```{r}
# cords for each individual
all_cords <- read.csv("coords_all.csv")

# cords for each population
cords <- read.csv("coords.csv")
```



# update the genlight object with population information
```{r}
pop(paxcan_gl) <- all_cords$NAME_OF_POPULATION[match(indNames(paxcan_gl), all_cords$INDIVIDUALS)]
```


# define colors for 14 pops
```{r}
# Define once, at the top, before any plot chunk
my.colors1 <- c("dodgerblue3", 
                "green", 
                "burlywood4", 
                "orange", 
                "yellow",
                "darkviolet", 
                "black", "tomato", "gold", "chocolate",
                "plum", "steelblue", 
                "forestgreen", "turquoise")
```


# Summarize data quality and diversity
```{r}
#glSum(paxcan_gl)    # numbers of loci & indivs
#glNA(paxcan_gl)     # missingness per sample
#glMean(paxcan_gl)   # allele frequencies per locus
#glVar(paxcan_gl)    # locus variances
```



# PCoA analyses
```{r}
# construct a dissimilarity matrix for the genetic data
D <- bitwise.dist(paxcan_gl, percent = FALSE, euclidean = TRUE, threads = 2)
```



## Run PCoA
```{r}
pcoa_res <- pcoa(D) # classical metric scaling

# extract scores for individuals
coords   <- as.data.frame(pcoa_res$vectors)

# add population factor
coords$pop <- pop(paxcan_gl)       

# percent variation explained by each axis
var_exp  <- 100 * pcoa_res$values$Relative_eig
head(var_exp)  # view first few axes
```


## confirm with dartR
```{r}
pcoa_res1 <- gl.pcoa(paxcan_gl, nfactors = 5, correction = "none", verbose = 3)
# same information is inside pcoa_res$eig (eigenvalues) and pcoa_res$scores

gl.pcoa.plot(pcoa_res1, paxcan_gl, xaxis = 1, yaxis = 2)
```


## visualize the results
```{r visualize PCoA results}
# MAKE A 2D PLOT WITH GGPLOT
ggplot(coords, aes(Axis.1, Axis.2, 
                   shape = pop, 
                   colour = pop)) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13)) +
  geom_point(size = 3) +
  labs(x = sprintf("PCo1 (%.1f%%)", var_exp[1]),
       y = sprintf("PCo2 (%.1f%%)", var_exp[2])) +
  theme_classic()

# Create interactive 3D PCoA plot with plotly
pcoa_3d <- plot_ly(coords, 
                   x = ~Axis.1, 
                   y = ~Axis.2, 
                   z = ~Axis.3,
                   color = ~pop,
                   colors = my.colors1,
                   type = "scatter3d",
                   mode = "markers",
                   marker = list(size = 5, 
                                 opacity = 0.8),
                   text = ~paste("Population:", pop),
                   hoverinfo = "text") %>%
  layout(
    title = "Interactive 3D PCoA of Population Structure",
    scene = list(
      xaxis = list(title = paste0("PCo1 (", round(var_exp[1],1), "%)")),
      yaxis = list(title = paste0("PCo2 (", round(var_exp[2],1), "%)")),
      zaxis = list(title = paste0("PCo3 (", round(var_exp[3],1), "%)"))
    )
  )

pcoa_3d
# looks like powers hollow, knobbs mountain and saint clairesville offer the most distinguishing featuures as the separate form the main cluster.  

# Save as HTML widget (interactive)
htmlwidgets::saveWidget(pcoa_3d, "pcoa_3d_interactive.html")
```



# PCA Analyses
```{r}
# Run PCA 
pca1 <- glPca(paxcan_gl, nf = 60) # let's retain the first 60 PCs

# extract scores
scores <- as.data.frame(pca1$scores)

# assign population information to scores
scores$pop <- pop(paxcan_gl)

# compute PCA-specific variance first, then use it
pca_var_exp <- 100 * pca1$eig / sum(pca1$eig)

head(pca_var_exp)
```



## visualize PCA in 2D and 3D
```{r pca visuals}
# MAKE A 2D PLOT WITH GGPLOT
pca_2d <- ggplot(scores, aes(PC1, PC2, 
                             shape = pop, 
                             colour = pop)) +
   scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13)) +
  geom_point(size = 3) +
  labs(x = sprintf("PCA1 (%.1f%%)", pca_var_exp[1]),
       y = sprintf("PCA2 (%.1f%%)", pca_var_exp[2])) +
  theme_classic()

# Save pca at 300 dpi
ggsave(
  filename = "pca.jpeg",
  plot = pca_2d,
  width = 8,
  height = 8,
  units = "in",
  dpi = 300,
  bg = "white"
)

# Create Word document and insert image
doc <- read_docx()

doc <- body_add_par(doc, "PCA Results", 
                    style = "heading 1")
doc <- body_add_img(
  doc,
  src = "pca.jpeg",
  width = 8,
  height = 8
)

print(doc, target = "pca_results_figure_6.docx")

# Create interactive 3D PCA plot with plotly
pca_3d <- plot_ly(scores, 
                  x = ~PC1, 
                  y = ~PC2, 
                  z = ~PC3,
                  color = ~pop,
                  colors = my.colors1,
                  type = "scatter3d",
                  mode = "markers",
                  marker = list(size = 5, opacity = 0.8),
                  text = ~paste("Population:", pop),
                  hoverinfo = "text") %>%
  layout(
    title = "Interactive 3D PCA of Population Structure",
    scene = list(
      xaxis = list(title = paste0("PC1 (", round(100*pca1$eig[1]/sum(pca1$eig),1), "%)")),
      yaxis = list(title = paste0("PC2 (", round(100*pca1$eig[2]/sum(pca1$eig),1), "%)")),
      zaxis = list(title = paste0("PC3 (", round(100*pca1$eig[3]/sum(pca1$eig),1), "%)"))
    )
  )

pca_2d
pca_3d

# Same results as that of the PCA.
# Save as HTML widget (interactive)
htmlwidgets::saveWidget(pca_3d, "pca_3d_interactive.html")
```



# DAPC ANALYSES
## run dapc based of our knowledge from the PCA above
```{r}
paxcan.dapc <- dapc(paxcan_gl, n.pca = 3, n.da = 2)
```



## plot the dapc
```{r}
# We would like our points to be shaped so we will need to modify the ggDAPC code to help us get it right. So here goes nothing:

# modified ggDAPC code
ggDapc_shaped <- function(dapc     = NULL,
                          k        = NULL,
                          x        = "LD1",
                          y        = "LD2",
                          cstar    = TRUE,
                          ellipse  = TRUE,
                          labels   = TRUE,
                         pt.size  = 3) {
  # internal data shaper (unchanged from ggDAPC source)
  ggDataShaper <- function(x, 
                           k = NULL) {
    if (is.null(k)) k <- x$grp
    da.coord  <- data.frame(x$ind.coord, 
                            grp = x$grp) %>% 
      as_tibble()
    grp.coord <- data.frame(x$grp.coord, 
                            grp = rownames(x$grp.coord))
    ind.coord <- left_join(da.coord, 
                           grp.coord, 
                           by = "grp", 
                           suffix = c("", ".c"))
    ind.coord$grp <- k
    labels  <- data.frame(x$grp.coord, 
                        "Pop" = row.names(x$grp.coord))
    list(ind.coord = ind.coord, grp.labels = labels)
  }
# scatter with shape added 
  ggScatter <- function(ggdata, x, y, cstar, ellipse, grp.labels, pt.size) {
    out <- ggplot(ggdata$ind.coord,
                  aes_string(x = x, y = y,
                             color = "grp",
                             #shape added here
                            shape = "grp")) +
      geom_point(size = pt.size) +
      scale_color_scico_d(palette = "romaO", 
                          begin = 0.2, end = 0.8) +
      scale_fill_scico_d(palette  = "romaO", 
                         begin = 0.2, end = 0.8) +
      theme(legend.position = "none") +
      theme_classic()

    if (cstar) {
      out <- out + geom_segment(
        data = ggdata$ind.coord,
        aes_string(x    = x,  xend = paste0(x, ".c"),
                   y    = y,  yend = paste0(y, ".c")),
        inherit.aes = FALSE
      )
    }
    if (ellipse) out <- out + stat_ellipse(
  type = "norm",
  na.rm  = TRUE
)
   if (grp.labels) out <- out + ggrepel::geom_text_repel(
  data        = ggdata$grp.labels,
  aes_string(x = x, y = y, label = "Pop"),
  color       = "black", fontface = "bold", size = 4,
  inherit.aes = FALSE,
  max.overlaps = Inf,
  box.padding  = 0.5
)
    out
  }
# density fallback for single DA (unchanged) 
  ggDensity <- function(ggdata, x) {
    d <- density(unlist(ggdata$ind.coord[, x]))
    ggplot(ggdata$ind.coord,
           aes_string(x = x, 
                      groups = "grp", 
                      fill = "grp", 
                      colour = "grp")) +
      geom_density(alpha = 0.7) +
      xlim(min(d$x), max(d$x)) +
      labs(fill = x, color = x) +
      scale_color_scico_d(palette = "romaO", 
                          begin = 0.2, end = 0.8) +
      scale_fill_scico_d(palette  = "romaO", 
                         begin = 0.2, end = 0.8) +
      theme_classic()
  }
shaped_data <- ggDataShaper(dapc, k)

  if (dapc$n.da > 1)
    ggScatter(shaped_data, 
              x, 
              y, 
              cstar, 
              ellipse, 
              labels, 
              pt.size)
  else
    ggDensity(shaped_data, x)
}
```



## apply modified functions
```{r DAPC visuals}
p_scatter <- ggDapc_shaped(paxcan.dapc,
                           x       = "LD1",
                           y       = "LD2",
                           cstar   = TRUE,
                           ellipse = TRUE,
                           labels  = TRUE,
                           pt.size = 3)

p_scatter <- p_scatter +
  scale_shape_manual(
    values = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13))

p_scatter
```


## bug fixes for ggDAPC
```{r}
source("https://raw.githubusercontent.com/wilsonfrantine/ggDAPC/main/ggDAPC.R")

## found a couple bugs in the ggPCvar, ggDAplot, and ggCompoplot functions in ggDAPC. These should fix them. They would overwrite the broken versions of those functions
ggPCvar <- function(x) {
  data.frame(
    var     = cumsum(x$pca.eig) / sum(x$pca.eig) * 100,
    ret.pca = c(rep(TRUE,  ncol(x$tab)),                        # use ncol not length
                rep(FALSE, length(x$pca.eig) - ncol(x$tab))),
    pca     = seq_along(x$pca.eig)
  ) %>%
    ggplot(aes(y = var, x = pca, fill = ret.pca)) +
    geom_col(width = 1) +
    theme_classic() +
    scale_fill_scico_d(palette = "grayC", begin = 0.1, end = 0.2) +
    theme(legend.position = "none") +
    labs(y = "PC cum. var. %", x = "PC")
}

## same issue for ggDAplot. This should overwrite the main funntion
ggDAplot <- function(x) {
  data.frame(
    DA     = colnames(x$loadings),
    # subset to n.da values only
    values = x$eig[seq_len(ncol(x$loadings))],   
    fill   = "gray"
  ) %>%
    ggplot(aes(x = DA, y = values, fill = fill)) +
    geom_col() +
    theme_classic() +
    labs(y = "DA eigenvalues", x = "") +
    scale_fill_scico_d(palette = "grayC", begin = 0.2) +
    theme(legend.position = "none")
}

## same for making the compoplot
ggCompoplot <- function(dapc) {
  # create a clean row order, then derive both the posterior matrix and the population labels from that same ordered index so they stay aligned.
  grp_vec <- as.character(dapc$grp)
  grp_order <- order(grp_vec)
  posterior_sorted <- dapc$posterior[grp_order, 
                                   , drop = FALSE]
  grp_sorted <- grp_vec[grp_order]
  
  # make the block boundaries from grp_sorted
  runs <- rle(grp_sorted)
  ends <- cumsum(runs$lengths)
  starts <- ends - runs$lengths + 1
  centers <- (starts + ends) / 2

  #make a temporary dataframe to hold the sorted values
temp.data <- data.frame(
  value = runs$values,
  x = starts,
  xend = ends + 1,
  labels.pos = centers,
  y = -0.05,
  yend = -0.05
)

# make plot
  out <- posterior_sorted %>%
  as_tibble() %>%
  mutate(ind = row_number()) %>%
  pivot_longer(cols = -ind, names_to = "grp", values_to = "prob") %>%
  ggplot(aes(y = prob, x = ind, fill = grp)) +
  geom_col(position = "stack", width = 1) +
  scale_x_continuous(
    breaks = temp.data$labels.pos,
    labels = temp.data$value
  ) +
  labs(y = "membership probability", x = "", fill = "") +
  theme_classic() +
  scale_fill_scico_d(palette = "romaO", begin = 0.2, end = 0.8, aesthetics = "fill") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
  return(out)
}
```


## scree plots and compoplots + combined figure
```{r DAPC screes and combined}
p_pcvar <- ggPCvar(paxcan.dapc)# cumulative PCA variance
p_da    <- ggDAplot(paxcan.dapc)   # DA eigenvalues
p_compo <- ggCompoplot(paxcan.dapc) # compo plot


# combine all four panels with patchwork
p_combined <- (p_scatter | (p_pcvar / p_da)) / p_compo +
  plot_layout(heights = c(3, 1))

p_combined

ggsave("dapc_scatter.jpeg",
       plot = p_combined,
       width = 10, height = 7,
       units = "in", dpi = 300, bg = "white")

ggsave("compoplot.jpeg",
       plot = p_compo,
       width = 10, height = 4,
       units = "in", dpi = 300, bg = "white")
```







# old code
```{r}
# dapc scatter plot
#Scatter plot (ggDapc replaces scatter()) 
p_scatter <- ggDapc(paxcan.dapc,
                    x= "LD1",
                    y= "LD2",
                    cstar= TRUE,    # lines from individual to group centroid
                    ellipse = TRUE, # confidence ellipses
                    labels  = TRUE)# group name labels


## save the dapc scatter plot as a jpeg at 300 dpi
jpeg(
  filename = "dapc.scatter.jpeg",
  width = 8,
  height = 8,
  units = "in",
  res = 300,
  bg = "white"
)
dev.off()

# Create Word document and insert image
doc <- read_docx()

doc <- body_add_par(doc, "DAPC Results Scatter Plot", 
                    style = "heading 1")
doc <- body_add_img(
  doc,
  src = "dapc.scatter.jpeg",
  width = 8,
  height = 8
)

print(doc, target = "dapc_results_figure_5.docx")


# compoplot
dapc.results <- as.data.frame(paxcan.dapc$posterior)
dapc.results$pop <- pop(paxcan_gl)
dapc.results$indNames <- rownames(dapc.results)

# make the dataframe longer for easy visualization
dapc.results <- pivot_longer(dapc.results, 
                             -c(pop, indNames))

# rename the columns
colnames(dapc.results) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")

# create the compoplot
p <- ggplot(dapc.results, 
            aes(x = Sample, 
                y = Posterior_membership_probability, 
                fill = Assigned_Pop)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = my.colors1) +
  facet_grid(~Original_Pop, scales = "free") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, 
                               hjust = 1, 
                               vjust = 1, 
                               size = 8)
  )
p

# Save compositional plot at 300 dpi
ggsave(
  filename = "compoplot.jpeg",
  plot = p,
  width = 8,
  height = 8,
  units = "in",
  dpi = 300,
  bg = "white"
)

# Create Word document and insert image
doc <- read_docx()

doc <- body_add_par(doc, "Paxican_compoplot", 
                    style = "heading 1")
doc <- body_add_img(
  doc,
  src = "compoplot.jpeg",
  width = 8,
  height = 8
)

print(doc, target = "supplemental_figure_1_compoplot.docx")
```


