---
title: "paxcan_basic_stats_analyses"
author: "Isaac Buabeng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## load the necessary libraires
```{r}
library(hierfstat)
library(gaston)
library(vcfR)
library(reshape2)
library(pegas)
library(gt)
library(dartR)
library(cowplot)
library(ggplot2)
library(officer)
library(tidyverse)
```



# load in the data
```{r}
vcf <- read.vcfR("14pops_biallelic_snps_miss0.4_thin10.newID.mac1.maf0.01.ldkb1r0.8.vcf")


# load in population data
# cords for each individual
all_cords <- read.csv("coords_all.csv")

# cords for each population
cords <- read.csv("coords.csv")
```


# calculate basic fstats
## convert the vcf file into a genind object for hierfstsat
```{r vcf2genind}
paxcan_genind <- vcfR2genind(vcf)
```



## convert the vcf object to a genlight object for dartR
```{r}
paxcan_gl <- vcfR2genlight(vcf)

# Ensure all required metadata and flags exist
paxcan_gl <- gl.compliance.check(paxcan_gl)   # adds missing flag skeletons

# Remove (or confirm absence of) monomorphic loci and set the flag
paxcan_gl <- gl.filter.monomorphs(paxcan_gl, verbose = 3)

# assign population names
pop(paxcan_gl) <- as.factor(all_cords$NAME_OF_POPULATION)

popNames(paxcan_gl)
```



## assign population information
```{r}
pop(paxcan_genind) <- as.factor(all_cords$NAME_OF_POPULATION)

popNames(paxcan_genind)
# perfect!
```



## convert the genind object into hierfstat format
```{r}
paxcan_hfstat <- genind2hierfstat(paxcan_genind, pop=NULL)
```



## calculate the basic fstats with hierfstat
```{r}
basicstat <- basic.stats(paxcan_hfstat, diploid=TRUE, digits=3)

# inspect the results
basicstat$overall

# save the results
save(basicstat, file="basicstat.RData")
```


## confirm with dartR
```{r}
gl.basic.stats(paxcan_gl)
# very similar to what we saw in hierfstat
```



## save the results into a dataframe
```{r}
overall_basic_stats <- as.data.frame(basicstat$overall)

# examine the table
overall_basic_stats
```



## Get bootstrap confidence values for Fis
```{r}
boot <- boot.ppfis(paxcan_hfstat, nboot=1000)
boot5 <-boot.ppfis(paxcan_hfstat, nboot=1000, quant=0.5)

boot
boot5
```



## Make a table for our bootstrapped observed and expected heterozygosity values
```{r}
# extract population names for constructing the dataframe
pop_order <- cords$NAME_OF_POPULATION

# use the names of the populations to rearrange the heterozygosity results
colnames(basicstat$Ho) <- pop_order
```



## construct the table for fis values
```{r}
Ho <- colMeans(basicstat$Ho,na.rm=T)
He <- colMeans(basicstat$Hs,na.rm=T)
Fis <- boot5$fis.ci$ll
new_popgen_stats <- cbind(Ho,He,Fis,boot$fis.ci)

new_popgen_stats
```



## save the resulst
```{r}
popgen_stats_noMD.csv <-write.table(new_popgen_stats, sep=",", dec=".", row.names=TRUE, col.names=TRUE)

save(new_popgen_stats, file="new_popgen_stats")
```



## calculate fst 
```{r}
paxcan.fst.mat <- genet.dist(paxcan_hfstat, method = "WC84")

# make the results into a matrix with column and rownames as population names
gindF.fst.mat.tri <- as.matrix(paxcan.fst.mat)

colnames(gindF.fst.mat.tri) <- pop_order

rownames(gindF.fst.mat.tri) <- pop_order

write.table(gindF.fst.mat.tri, file="paxcanFstWC84.pwfst", row.names=T,col.names=T)
```



## construct a table for the results
```{r}
melted <- melt(gindF.fst.mat.tri, na.rm = TRUE)
```


## look at results
```{r}
summary(melted$value)
```



## pairwise fst plot based on disjunctions
```{r}
# Calculate mean FST values and create ordered site list
mean_fst_by_pop <- melted %>%
  group_by(Var1) %>%
  summarise(mean_fst = mean(value, na.rm = TRUE)) %>%
  arrange(mean_fst)

ordered_sites <- mean_fst_by_pop$Var1

# Define populations by division
central_populations <- c(
  "Saint Clairesville", "Shawnee State Park", "Lutzville", 
  "Town Creek", "False Castle Rock", "Bluerock", "SELU", 
  "Smoke Hole", "The Pinnacle", "Brush Creek Falls1", 
  "Powers Hollow", "Knobbs Mountain"
)

cumberland_populations <- c("Carter Caves", "Sweet Lilly Ridge")

# Order each division by mean FST values
central_ordered <- ordered_sites[ordered_sites %in% central_populations]
cumberland_ordered <- ordered_sites[ordered_sites %in% cumberland_populations]

# Create final site order
site_order <- c(central_ordered, cumberland_ordered)

# Set factor levels
melted$Var1 <- factor(melted$Var1, levels = site_order)
melted$Var2 <- factor(melted$Var2, levels = site_order)

# Create heatmap 
combined <- ggplot(melted, aes(Var2, Var1, fill = value)) + 
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red", name = "FST") +
  labs(x = "Sampling Site", y = "Sampling Site") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 13)) +
  coord_fixed()

# Add division lines (adjust based on new ordering)
division_line_position <- length(central_ordered) + 0.5

combined <- combined +
  geom_vline(xintercept = division_line_position, linewidth = 1.2) +
  geom_hline(yintercept = division_line_position, linewidth = 1.2)

# Add group labels
# Cumulative block sizes
cum_central <- length(central_ordered) # last row of Central Appalachian block
cum_total   <- length(central_ordered) +               
               length(cumberland_ordered) # last row overall

# Division labels aligned to top edge 
combined <- combined +
  annotate("text",
           x = 0.9, # small left offset inside the facet
           y = cum_central,   # top edge of first block
           label    = "Central Appalachian\nMountains",
           hjust    = 0,    # center-align
           vjust    = 1,     # anchor to top of the row
           angle    = 0,
           size     = 4.5,
           fontface = "bold") +
  annotate("text",
           x = 0.9,
           y = cum_total, # top edge of second block (= last row)
           label = "Cumberland–Allegheny\nInterior Plateaus",
           hjust    = 0, # center-align
           vjust    = 1,
           angle    = 0,
           size     = 4.5,
           fontface = "bold") +
  theme(plot.margin = margin(t = 10, r = 10, b = 10, l = 50))

combined


# Save heatmap at 300 dpi
ggsave(
  filename = "fst_heatmap_300dpi.jpeg",
  plot = combined,
  width = 8,
  height = 8,
  units = "in",
  dpi = 300,
  bg = "white"
)

# Create Word document and insert image
doc <- read_docx()

doc <- body_add_par(doc, "FST Heatmap", 
                    style = "heading 1")
doc <- body_add_img(
  doc,
  src = "fst_heatmap_300dpi.jpeg",
  width = 8,
  height = 8
)

print(doc, target = "fst_heatmap_figure_4.docx")
```




## pairwise fst plot based on states
```{r}
### Compute a per-population mean FST
mean_fst <- melted %>%                         # melted has Var1, Var2, value
  group_by(Var1) %>%                           # one row per focal population
  summarise(mean_fst = mean(value, na.rm = TRUE)) %>% 
  arrange(mean_fst)                            # ascending order

ordered_all <- mean_fst$Var1                  # vector of names ranked by FST

### Define the five state groupings
pa   <- c("Saint Clairesville", "Shawnee State Park", "Lutzville")
md   <- c("Town Creek")
wv   <- c("False Castle Rock", "Bluerock", "Smoke Hole",
          "Brush Creek Falls1", "Powers Hollow")
va   <- c("The Pinnacle", "SELU", "Knobbs Mountain")
ky   <- c("Carter Caves", "Sweet Lilly Ridge")

### Re-order each state block by the ranked list
pa_order <- ordered_all[ordered_all %in% pa]
md_order <- ordered_all[ordered_all %in% md]
wv_order <- ordered_all[ordered_all %in% wv]
va_order <- ordered_all[ordered_all %in% va]
ky_order <- ordered_all[ordered_all %in% ky]

site_order1 <- c(pa_order, md_order, wv_order, va_order, ky_order)

### Set factor levels with the new order
melted$Var1 <- factor(melted$Var1, levels = site_order1)
melted$Var2 <- factor(melted$Var2, levels = site_order1)

## Relabel the states (the y coordinates are now midpoints of each block)
combined1 <- ggplot(melted, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red", name = "FST") +
  ggtitle(expression(atop("Pairwise FST, WC (1984), Combined",
                          atop(italic("N = 165"), "")))) +
  labs(x = "Sampling Site", y = "Sampling Site") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),
        axis.text.y = element_text(size = 13),
        plot.title  = element_text(size = 14, face = "bold"),
        plot.margin = margin(t = 10, r = 10, b = 10, l = 60)) +
  coord_fixed()

## Add state separators at the new boundaries
breaks <- c(length(pa_order),
            length(pa_order)+length(md_order),
            length(pa_order)+length(md_order)+length(wv_order),
            length(pa_order)+length(md_order)+length(wv_order)+length(va_order)) + 0.5

combined1 <- combined1 +
  geom_vline(xintercept = breaks, linewidth = 1.2) +
  geom_hline(yintercept = breaks, linewidth = 1.2)

## Position labels at the TOP EDGE of each state block
combined1 <- combined1 +
  annotate("text", x = 0.2, y = length(pa_order),                                              label = "Pennsylvania",   fontface = "bold", size = 4.5, hjust = 0, vjust = 1) +
  annotate("text", x = 0.2, y = length(pa_order) + length(md_order),                          label = "Maryland",       fontface = "bold", size = 4.5, hjust = 0, vjust = 1) +
  annotate("text", x = 0.2, y = length(pa_order) + length(md_order) + length(wv_order),       label = "West Virginia",  fontface = "bold", size = 4.5, hjust = 0, vjust = 1) +
  annotate("text", x = 0.2, y = length(pa_order) + length(md_order) + length(wv_order) + length(va_order), label = "Virginia", fontface = "bold", size = 4.5, hjust = 0, vjust = 1) +
  annotate("text", x = 0.2, y = length(site_order1),                                          label = "Kentucky",       fontface = "bold", size = 4.5, hjust = 0, vjust = 1)

combined1
```



## extract numbers for hetrozygosity
```{r}
paxcanHet <- summary(paxcan_genind)
paxcanHet

save(paxcanHet,file="paxcanHet.Rdata")

names(paxcanHet)
```


## extract plots for observed and expect heterozygosity at each locus
```{r}
plot(paxcanHet$Hobs, xlab="Locus number", ylab="Observed heterozygosity", main="Observed heterozygosity per locus")

plot(paxcanHet$Hobs, paxcanHet$Hexp, xlab="Observed heterozygosity", ylab="Expected heterozygosity", main="He as a function of Ho, per locus")


tiff(filename="paxcan_Hob_plot.tiff", width = 4, height = 4, units = 'in', res = 300)

tiff(filename="paxcan_Ho_He_plot.tiff", width = 4, height = 4, units = 'in', res = 300)

dev.off()
```



# Perform Barlett Test to test for the homogeneity of variances
```{r}
bartlett.test(list(paxcanHet$Hexp, paxcanHet$Hobs))

# so the observed heterozygosity is higher and statistically significant at that.
```


## Now, lets separate the pops and repeat Bartlett for each.
```{r}
temp_seppop <- seppop(paxcan_genind)
```


## Let's make a list of the populations
```{r}
PH <- temp_seppop$`Powers Hollow`
SC <- temp_seppop$`Saint Clairesville`
SELU <- temp_seppop$SELU
SP <- temp_seppop$`Shawnee State Park`
KM <- temp_seppop$`Knobbs Mountain`
BR <- temp_seppop$`Blue Rock`
BF <- temp_seppop$`Brush Creek Falls 1`
TC <- temp_seppop$`Town Creek`
SH <- temp_seppop$`Smoke Hole`
PIN <- temp_seppop$`The Pinnacle`
SLR <- temp_seppop$`Sweet Lilly Ridge`
CC <- temp_seppop$`Carter Caves`
FCR <- temp_seppop$`False Castle Rock`
LE <- temp_seppop$Lutzville
```



## Run a bartlett test for each population
### PH
```{r}
PHhet <- summary(PH)

bartlett.test(list(PHhet$Hexp, PHhet$Hobs))
```


### SELU
```{r}
SELUhet <- summary(SELU)

bartlett.test(list(SELUhet$Hexp, SELUhet$Hobs))
```



### SC
```{r}
SChet <- summary(SC)

bartlett.test(list(SChet$Hexp, SChet$Hobs))
```



### SP
```{r}
SPhet <- summary(SP)

bartlett.test(list(SPhet$Hexp, SPhet$Hobs))
```



### KM
```{r}
KMhet <- summary(KM)

bartlett.test(list(KMhet$Hexp, KMhet$Hobs))
```



### BR
```{r}
BRhet <- summary(BR)

bartlett.test(list(BRhet$Hexp, BRhet$Hobs))
```



### BF
```{r}
BFhet <- summary(BF)

bartlett.test(list(BFhet$Hexp, BFhet$Hobs))
```



### TC
```{r}
TChet <- summary(TC)

bartlett.test(list(TChet$Hexp, TChet$Hobs))
```



### SH
```{r}
SHhet <- summary(SH)

bartlett.test(list(SHhet$Hexp, SHhet$Hobs))
```



### PIN
```{r}
PINhet <- summary(PIN)

bartlett.test(list(PINhet$Hexp, PINhet$Hobs))
```



### SLR
```{r}
SLRhet <- summary(SLR)

bartlett.test(list(SLRhet$Hexp, SLRhet$Hobs))
```


### CC
```{r}
CChet <- summary(CC)

bartlett.test(list(CChet$Hexp, CChet$Hobs))
```



### FCR
```{r}
FCRhet <- summary(FCR)

bartlett.test(list(FCRhet$Hexp, FCRhet$Hobs))
```



### LE
```{r}
LEhet <- summary(LE)

bartlett.test(list(LEhet$Hexp,LEhet$Hobs))
```
