---
title: "paxcan_clonality_analyses"
author: "Isaac Buabeng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## load the necessary libraries
```{r}
library(cowplot)
library(vcfR)
library(poppr)
library(adegenet)
library(ggplot2)
library(reshape2)
library(pheatmap)
library(tidyverse)
```



# Import LD-pruned and non-LD pruned data
```{r}
#non-ld pruned
vcf_data <- read.vcfR("14pops_biallelic_snps_miss0.4_thin10.newID.mac1.maf0.01.vcf")

#ld pruned
vcf_data1 <- read.vcfR("14pops_biallelic_snps_miss0.4_thin10.newID.mac1.maf0.01.ldkb1r0.8.vcf")
```


# Add population information
```{r}
coords <- read.csv("coords_all.csv")
snp_genind <- vcfR2genind(vcf_data1, return.alleles = TRUE)
pop(snp_genind) <- coords$NAME_OF_POPULATION
```



# let's check if there are signals of clonality using the non-pruned data
```{r}
# convert vcf to genlight object
snp_genlight <- vcfR2genlight(vcf_data)

# Add population information
pop(snp_genlight) <- coords$NAME_OF_POPULATION
snp_snpclone <- as.snpclone(snp_genlight)

# Random sampling approach (recommended for large SNP datasets)
ia_results <- samp.ia(snp_snpclone, 
                      n.snp = 100,    # Sample 100 SNPs at a time
                      reps = 1000)    # Repeat 1000 times

# View results
plot(ia_results, main = "Distribution of r̄d Values")
summary(ia_results)
```



# Population-Specific IA Testing
```{r}
# Test each population separately
ia_by_pop <- lapply(seppop(snp_snpclone), function(x) {
  samp.ia(x, n.snp = 100, reps = 500)
})

# Summarize results
ia_summary <- data.frame(
  Population = names(ia_by_pop),
  Mean_rbarD = sapply(ia_by_pop, mean),
  Median_rbarD = sapply(ia_by_pop, median),
  P_value = sapply(ia_by_pop, function(x) sum(x >= 0) / length(x))
)

print(ia_summary)

# so we have strong signaling for sexual reproduction instead of clonality. our populations are predominantly sexual with little to no clonal reproduction based on the Index of Association test. The rbarD values, while positive, are not significantly different from what you'd expect under random mating.
```


# clone counting
# Convert to genclone and count clones
```{r}
snp_genclone <- as.genclone(snp_genind)
mlg.filter(snp_genclone, distance = bitwise.dist) <- 0.01
```



# Get clone counts per population
```{r}
clone_counts <- rowSums(mlg.table(snp_genclone) > 0)
print(clone_counts)
```


# validation of clonality results
```{r}
# Compare multiple thresholds as done in literature
thresholds <- c(0.005, 0.01, 0.02)
clone_counts <- sapply(thresholds, function(t) {
  mlg.filter(snp_genclone, distance = bitwise.dist) <- t
  rowSums(mlg.table(snp_genclone) > 0)
})

# Examine genetic distance distribution (as recommended in literature)
distances <- bitwise.dist(snp_genclone)
hist(as.vector(distances), breaks = 50, main = "Genetic Distance Distribution")
```



## visualization
```{r}
# After calculating clone counts
clone_results <- data.frame(
  Population = names(clone_counts),
  Clone_Count = clone_counts,
  Sample_Size = table(pop(snp_genclone))
)

# Basic bar chart
ggplot(clone_results, aes(x = Population, y = Clone_Count, fill = Population)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  scale_fill_viridis_d() +  # Nice color palette
  labs(title = "Number of Multilocus Genotypes per Population",
       x = "Population", y = "Number of MLGs") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +  # Remove legend since x-axis shows populations
  coord_cartesian(ylim = c(0, max(clone_results$Clone_Count) * 1.1))  # Add some space at top
```



# counting how many true clones there are
```{r}
# Check for actual clones
mlg_table <- mlg.table(snp_genclone)
repeated_mlgs <- which(rowSums(mlg_table > 1) > 0)
clone_frequency <- mlg_table[repeated_mlgs, ]

print(clone_frequency)


# Convert to long format for ggplot2
clone_long <- melt(as.matrix(clone_frequency), 
                   varnames = c("Population", "MLG"), 
                   value.name = "Count")

# Remove zeros for cleaner visualization
clone_long <- clone_long[clone_long$Count > 0, ]

pheatmap(clone_frequency,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("white", "darkblue"))(50),
         main = "MLG Frequency Across Populations",
         fontsize_row = 8,
         fontsize_col = 10,
         display_numbers = TRUE)  # Shows actual counts in cells
```



# addressing population size differences
```{r}
# Check sample sizes per population
pop_sizes <- table(pop(snp_genclone))
print(pop_sizes)

# Identify problematic populations
small_pops <- pop_sizes[pop_sizes < 8]
if(length(small_pops) > 0) {
  cat("Populations with <8 individuals:\n")
  print(small_pops)
}
```



# remove populations with less than 8 individuals and re-run the IA analyses
## re-load the data first and rerun the analyses
```{r}
# You'll need to re-import and filter your VCF data
vcf_data_filtered <- read.vcfR("14pops_biallelic_snps_miss0.4_thin10.newID.mac1.maf0.01.vcf")
snp_genlight_all <- vcfR2genlight(vcf_data_filtered)

# Add population information
coords <- read.csv("coords_all.csv")
pop(snp_genlight_all) <- coords$NAME_OF_POPULATION

# Filter to reliable populations only
reliable_pops <- names(pop_sizes)[pop_sizes >= 8]
snp_genlight_filtered <- snp_genlight_all[pop(snp_genlight_all) %in% reliable_pops]

# Convert to snpclone
snp_snpclone_filtered <- as.snpclone(snp_genlight_filtered)

# Now run global IA analysis
new_ia_results <- samp.ia(snp_snpclone_filtered, n.snp = 100, reps = 1000)

plot(new_ia_results)
```





## Rerun Population-Specific IA Testing
```{r}
# Test each population separately
ia_by_pop1 <- lapply(seppop(snp_snpclone_filtered), function(x) {
  samp.ia(x, n.snp = 100, reps = 500)
})

# Summarize results
ia_summary1 <- data.frame(
  Population = names(ia_by_pop1),
  Mean_rbarD = sapply(ia_by_pop1, mean),
  Median_rbarD = sapply(ia_by_pop1, median),
  P_value = sapply(ia_by_pop1, function(x) sum(x >= 0) / length(x))
)

print(ia_summary1)

# so we still have strong signaling for sexual reproduction instead of clonality. our populations are predominantly sexual with little to no clonal reproduction based on the Index of Association test. The rbarD values, while positive, are not significantly different from what you'd expect under random mating.
```
