---
title: "paxcan_snmf_analyses"
author: "Isaac Buabeng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Description of the analyses: The analyses performed in this script were performed following the instructions and suggestions made in http://membres-timc.imag.fr/Olivier.Francois/tutoRstructure.pdf


## Load the necessary libraries
```{r}
library(vcfR)
library(LEA)
library(ggplot2)
library(ggspatial)
library(sf)
library(rnaturalearth)    # for easy sf basemaps
library(scatterpie)
library(scales)
library(ggnewscale)
library(tidyverse)
```



# load data and set genotype
```{r}
# load data
paxcan.geno <- vcf2geno("14pops_biallelic_snps_miss0.4_thin10.newID.mac1.maf0.01.ldkb1r0.8.vcf")

#set genotype
genotype = paxcan.geno
```


# run snmf
```{r}
paxcan_snmf = snmf(genotype, K = 1:14, entropy = T, ploidy=2, project = "new")
```



# Running 1000 iterations on the initial run
```{r}
set.seed(12334)
paxcan_snmf_reps = snmf(genotype, 
                        K = 1:14, 
                        CPU = 4, 
                        entropy = T, 
                        ploidy=2, 
                        project = "new", 
                        iterations = 1000, 
                        repetitions = 10)
```



# find the number of clusters
```{r}
set.seed(12334)
# make an initial plot of the reps
plot(paxcan_snmf_reps)

# get the cross-entropy value for each run
ce_min <- sapply(1:14, function(k) min(cross.entropy(paxcan_snmf_reps, K = k)))

#get the change in values between each k
delta_ce <- diff(ce_min)

# plot the distribution of changes across the different Ks
plot(2:14, delta_ce, 
     type = "b", 
     pch = 19,
     col = "blue",
     xlab = "K (Number of Ancestral Populations)",
     ylab = "Change in Cross-Entropy (ΔCE)",
     main = "Cross-Entropy Slope Analysis")


# Threshold of -0.01 means "improvement less than 0.01"
abline(h = -0.01, col = "red", lty = 2, lwd = 3) # heuristic cutoff

# Add text annotation
text(x = 8, y = -0.01, 
     labels = "Diminishing returns threshold (-0.01)", 
     pos = 3, col = "red", font = 2)

# Identify elbow point where improvement becomes small
elbow_k <- which(delta_ce >= -0.01)[1] + 1  # First K where improvement < 0.01
if(!is.na(elbow_k)) {
  abline(v = elbow_k, col = "green", lty = 3, lwd = 2)
  text(x = elbow_k, y = min(delta_ce), 
       labels = paste("Elbow at K =", elbow_k), 
       pos = 4, col = "green", font = 2)

}
# looks like we have a cut off at k=11
```


# Now let's plot the ideal Ks
## verifying names
```{r}
# Read VCF file 
vcf <- read.vcfR("14pops_biallelic_snps_miss0.4_thin10.newID.mac1.maf0.01.ldkb1r0.8.vcf")

# Extract sample names from VCF header
vcf_sample_names <- colnames(vcf@gt)[-1]  # Remove first column (FORMAT)
print("Sample names from VCF:")
print(vcf_sample_names)

# Read your sample names file
sample_names_file <- readLines("C:/Users/User/Documents/paxcan_project/sample_names")
print("Sample names from file:")
print(sample_names_file)

# Check if they match in order
if(identical(vcf_sample_names, sample_names_file)) {
  print("Sample names match perfectly!")
} else {
  print("Sample names don't match or are in different order")
  print("Checking differences...")
}
```



# visualize k=14 for the 14 populations in the initial analyses
```{r}
# 1. Determine optimal run for K = 14
best14 <- which.min(cross.entropy(paxcan_snmf_reps, K = 14))

# 2. Custom colour palette
my.colors1 <- c("dodgerblue3", "green", "burlywood4", "orange", "yellow",
                "darkviolet", "black", "tomato", "gold", "chocolate", 
                "plum", "steelblue", "forestgreen", "turquoise")

# 3. Extract Q-matrix for that run
q_matrix <- Q(paxcan_snmf_reps, K = 14, run = best14)

## 4. Plot using LEA's barchart function with sample labels
barchart(paxcan_snmf_reps, K = 14, run = best14,
         border = NA, 
         space = 0,
         col = my.colors1,
         xlab = "Individuals",
         ylab = "Ancestry proportions",
         main = "Ancestry matrix",
         lab = TRUE,              # Provide sample names here
         sort.by.Q = TRUE) -> bp         # arrange according structuring

# The barchart function should handle the labeling automatically

axis(1, at = 1:length(bp$order), 
      labels = bp$order, las = 3, 
      cex.axis = .4)


# Aggregate by population for cleaner distribution visualization
# prepare a table for plotting
pop_distribution <- all_cords %>% # all cords is from the dapc analyses and contains cords for each of the populations and individuals
  bind_cols(as.data.frame(q_matrix)) %>%
  group_by(NAME_OF_POPULATION, lon., lat.) %>%
  summarise(across(V1:V14, mean), .groups = 'drop')

pop_distribution$pie_radius <- diff(range(pop_distribution$lat.)) * 0.07

# get the base maps
world_sf  <- ne_countries(scale = "medium", returnclass = "sf") 
states_sf <- ne_states(country = "United States of America", returnclass = "sf")

# Assign a distinct color per population
pop_cols <- hue_pal()(nrow(pop_distribution))
names(pop_cols) <- pop_distribution$NAME_OF_POPULATION

ggplot() +
  # Base maps
geom_sf(data = world_sf, fill = "#f0f0f0", color = NA) +
  geom_sf(data = states_sf, fill = NA, color = "gray50", size = 0.3) +
  
  # Pie slices (fill legend: V1–V14)
  geom_scatterpie(
    data = pop_distribution,
    aes(x = lon., y = lat., r = pie_radius, group = NAME_OF_POPULATION),
    cols = paste0("V", 1:14),
    color = "white", alpha = 0.85,
    legend_name = "Ancestralproportions"
  ) +
  
  # Switch to new fill/color scale for population colors
  new_scale_color() +
  
  # points colored by population
  geom_point(
    data = pop_distribution,
    aes(x = lon., y = lat., color = NAME_OF_POPULATION),
    size = 3, shape = 21, stroke = 1
  ) +
  
  scale_color_manual(
    name = "Population",
    values = pop_cols
  ) +
  
  # Map elements
  coord_sf(
    xlim = range(pop_distribution$lon.) + c(-2, 2),
    ylim = range(pop_distribution$lat.) + c(-2, 2),
    expand = FALSE
  ) +
  annotation_scale(location = "bl", width_hint = 0.2) +
  annotation_north_arrow(
    location = "bl", which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +
  
  # Pie size legend
  geom_scatterpie_legend(
    pop_distribution$pie_radius,
    x = min(pop_distribution$lon.),
    y = min(pop_distribution$lat.) - 1,
    n = 3,
    labeller = function(x) round(pi * x^2, 2)
  ) +
  
  # Styling
  theme_void() +
  theme(
    legend.position.inside = c(0.85, 0.25),
    legend.background = element_rect(fill = "white", color = NA)
  ) +
  ggtitle("Population Admixture (K = 14)")


# save the map
ggsave("admixture_map_k14.png", 
       width = 12, height = 8, dpi = 300)
```



# visualize k=12 for our best K from the cross entropy analyses
```{r}
set.seed(123)
# Determine optimal run for K = 11
best12 <- which.min(cross.entropy(paxcan_snmf_reps, K = 12))

# Custom colour palette
my.colors2 <- c("dodgerblue3", "green", "burlywood4", "orange", "yellow",
                "darkviolet", "black", "tomato", "gold", "chocolate", "plum", "turquoise")

# Extract Q-matrix for that run
q_matrix1 <- Q(paxcan_snmf_reps, K = 12, run = best12)


# Convert Q-matrix into long format
df <- as.data.frame(q_matrix1)
df$sample <- rownames(df)
df_long <- pivot_longer(df, cols = starts_with("V"), 
                        names_to = "Component", values_to = "Proportion")

# Ensure sample ordering matches bp1$order
df_long$sample <- factor(df_long$sample, levels = bp1$order)

ggplot(df_long, aes(x = sample, y = Proportion, fill = Component)) +
  geom_col() +
  scale_fill_manual(values = my.colors2, name = "Ancestral Component") +
  labs(x = "Individuals", y = "Ancestry proportions") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 6))

```




# visualize k=2 for our best K from the cross entropy analyses
```{r}
# Determine optimal run for K = 2
best2 <- which.min(cross.entropy(paxcan_snmf_reps, K = 2))

# Custom colour palette
my.colors3 <- c("dodgerblue3", "green")

# Extract Q-matrix for that run
q_matrix2 <- Q(paxcan_snmf_reps, K = 2, run = best2)

## Plot using LEA's barchart function with sample labels
barchart(paxcan_snmf_reps, K = 2, run = best2,
         border = NA, 
         space = 0,
         col = my.colors3,
         xlab = "Individuals",
         ylab = "Ancestry proportions",
         main = "Ancestry matrix",
         lab = TRUE,
         sort.by.Q = TRUE) -> bp2         # arrange according structuring

axis(1, at = 1:length(bp2$order), 
      labels = bp2$order, las = 3, 
      cex.axis = .4)

# NO CLEAR SPLIT BETWEEN THE TWO DISJUNCTIONS
```