---
title: "paxcan_IBD_AMOVA_analyses"
author: "Isaac Buabeng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## load the necessary libraries
```{r}
library(adegenet)
library(pegas)
library(vcfR)
library(dartR)
library(poppr)
```



# load in the data
```{r read in vcf with vcfR}
vcf <- read.vcfR("14pops_biallelic_snps_miss0.4_thin10.newID.mac1.maf0.01.ldkb1r0.8.vcf")
```



# convert the vcf object to a genlight object
```{r}
paxcan_gl <- vcfR2genlight(vcf)

# Ensure all required metadata and flags exist
paxcan_gl <- gl.compliance.check(paxcan_gl)   # adds missing flag skeletons

# Remove (or confirm absence of) monomorphic loci and set the flag
paxcan_gl <- gl.filter.monomorphs(paxcan_gl, verbose = 3)
```



# load in population data
```{r}
# cords for each individual
all_cords <- read.csv("coords_all.csv")

# cords for each population
cords <- read.csv("coords.csv")
```



# assign population names
```{r}
pop(paxcan_gl) <- as.factor(all_cords$NAME_OF_POPULATION)

popNames(paxcan_gl)
```

 
# generate arguments for mantel test for IBD
```{r generate DGEO and DGEN}
Dgen <- gl.dist.pop(paxcan_gl)
#Dgen

Dgeo <- dist(cords)
#Dgeo
```


# perform the mantel test
```{r}
paxcan_ibd <- mantel.randtest(Dgen, Dgeo)

plot(paxcan_ibd)

#Monte-Carlo test
#Call: mantel.randtest(m1 = Dgen, m2 = Dgeo)

#Observation: 0.06225107 

#Based on 999 replicates
#Simulated p-value: 0.386 
#Alternative hypothesis: greater 

#     Std.Obs  Expectation     Variance 
# 0.365965692 -0.001738827  0.030573287 
  
# there is no correlation between genetic distance and geographic distance




## save the ibd plot as a jpeg at 300 dpi
jpeg(
  filename = "IBD.jpeg",
  width = 8,
  height = 8,
  units = "in",
  res = 300,
  bg = "white"
)

## draw plot into the file
# confirming with dartr
paxcan_ibd1 <- gl.ibd(x = NULL, Dgen = Dgen, 
                      Dgeo = Dgeo, 
                      permutations = 999)

dev.off()

# Create Word document and insert image
doc <- read_docx()

doc <- body_add_par(doc, "IBD Results", 
                    style = "heading 1")
doc <- body_add_img(
  doc,
  src = "IBD.jpeg",
  width = 8,
  height = 8
)

print(doc, target = "IBD_results_figure_7.docx")

#Analysis performed using provided genetic and Euclidean distance matrices. If a genlight object is provided, it is ignored.
#  Coordinates used from: Dgeo provided. 
#  Transformation of Dgeo: Dgeo 
#  Genetic distance: Dgen provided 
#  Tranformation of Dgen:  Dgen 

#Mantel statistic based on Pearson's product-moment correlation 

#Call:
#vegan::mantel(xdis = Dgen, ydis = Dgeo, permutations = permutations,      na.rm = TRUE) 

#Mantel statistic r: 0.06225 
#      Significance: 0.402 

#Upper quantiles of permutations (null model):
#  90%   95% 97.5%   99% 
#0.223 0.266 0.300 0.335 
#Permutation: free
#Number of permutations: 999
#Completed: gl.ibd
```



# Perform Analyses of Molecular Variance (AMOVA)
## assign the strata file for proper stratification analyses
```{r}
paxcan_strata <- read.csv("strata_file.csv")

paxcan_strata
```



# assign strata the genlight object
```{r}
strata(paxcan_gl) <- paxcan_strata

paxcan_gl$strata
```



## perform amova on the population basis
```{r}
paxcan_amova_pop <- poppr.amova(paxcan_gl, hier= ~POPULATION)

paxcan_amova_pop
```



## perform amova on a statewide level
```{r}
paxcan_amova_state <- poppr.amova(paxcan_gl, hier= ~LOCATION)

paxcan_amova_state
```



## perform amova on discontinuity level
```{r}
paxcan_amova_region <- poppr.amova(paxcan_gl, hier= ~REGION)

paxcan_amova_region
```

