---
title: "paxcan_pca_pcoa_dapc_analyses"
author: "Isaac Buabeng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## load the necessary libraries
```{r}
library(adegenet)
library(pegas)
library(vcfR)
library(ape)
library(dartR)
library(plotly)
library(poppr) 
library(ggplot2)
library(tidyverse)
```



# load in and convert the data into a genlight object
```{r}
vcf <- read.vcfR("C:/Users/User/Documents/paxcan_project/14pops_biallelic_snps_miss0.4_thin10.newID.mac1.maf0.01.ldkb1r0.8.vcf")

# convert the vcf file into a genlight object
paxcan_gl <- vcfR2genlight(vcf)
```



# Filter loci and individuals
```{r}
# Ensure all required metadata and flags exist
paxcan_gl <- gl.compliance.check(paxcan_gl)   # adds missing flag skeletons

# Remove (or confirm absence of) monomorphic loci and set the flag
paxcan_gl <- gl.filter.monomorphs(paxcan_gl, verbose = 3)
```



# load in population data
```{r}
# cords for each individual
all_cords <- read.csv("coords_all.csv")

# cords for each population
cords <- read.csv("coords.csv")
```



# update the genlight object with population information
```{r}
pop(paxcan_gl) <- all_cords$NAME_OF_POPULATION[match(indNames(paxcan_gl), all_cords$INDIVIDUALS)]
```


# Summarize data quality and diversity
```{r}
glSum(paxcan_gl)    # numbers of loci & indivs
glNA(paxcan_gl)     # missingness per sample
glMean(paxcan_gl)   # allele frequencies per locus
glVar(paxcan_gl)    # locus variances
```


# PCoA analyses
```{r}
# construct a dissimilarity matrix for the genetic data
D <- bitwise.dist(paxcan_gl, percent = FALSE, euclidean = TRUE, threads = 2)
```



## Run PCoA
```{r}
pcoa_res <- pcoa(D) # classical metric scaling

# extract scores for individuals
coords   <- as.data.frame(pcoa_res$vectors)

# add population factor
coords$pop <- pop(paxcan_gl)       

# percent variation explained by each axis
var_exp  <- 100 * pcoa_res$values$Relative_eig
head(var_exp)                      # view first few axes
```


## confirm with dartR
```{r}
pcoa_res1 <- gl.pcoa(D, nfactors = 5, correction = "none", verbose = 3)
# same information is inside pcoa_res$eig (eigenvalues) and pcoa_res$scores

scatter(pcoa_res1)
```


## visualize the results
```{r visualize PCoA results}
# MAKE A 2D PLOT WITH GGPLOT
ggplot(coords, aes(Axis.1, Axis.2, colour = pop)) +
  geom_point(size = 3) +
  labs(x = sprintf("PCo1 (%.1f%%)", var_exp[1]),
       y = sprintf("PCo2 (%.1f%%)", var_exp[2])) +
  theme_classic()

# make a custom set of colors for the 14 pops
my.colors1 <- c("dodgerblue3", "green", "burlywood4", "orange", "yellow",
                "darkviolet", "black", "tomato", "gold", "chocolate", 
                "plum", "steelblue", "forestgreen", "turquoise")

# Create interactive 3D PCoA plot with plotly
pcoa_3d <- plot_ly(coords, 
                   x = ~Axis.1, 
                   y = ~Axis.2, 
                   z = ~Axis.3,
                   color = ~pop,
                   colors = my.colors1,
                   type = "scatter3d",
                   mode = "markers",
                   marker = list(size = 5, opacity = 0.8),
                   text = ~paste("Population:", pop),
                   hoverinfo = "text") %>%
  layout(
    title = "Interactive 3D PCoA of Population Structure",
    scene = list(
      xaxis = list(title = paste0("PCo1 (", round(var_exp[1],1), "%)")),
      yaxis = list(title = paste0("PCo2 (", round(var_exp[2],1), "%)")),
      zaxis = list(title = paste0("PCo3 (", round(var_exp[3],1), "%)"))
    )
  )

pcoa_3d
# looks like powers hollow, knobbs mountain and saint clairesville offer the most distinguishing featuures as the separate form the main cluster.  

# Save as HTML widget (interactive)
htmlwidgets::saveWidget(pcoa_3d, "pcoa_3d_interactive.html")
```



# PCA Analyses
```{r}
# Run PCA 
pca1 <- glPca(paxcan_gl, nf = 60) # let's retain the first 60 PCs

# extract scores
scores <- as.data.frame(pca1$scores)

# assign population information to scores
scores$pop <- pop(paxcan_gl)
```



## visualize PCA in 2D and 3D
```{r}
# MAKE A 2D PLOT WITH GGPLOT
ggplot(scores, aes(PC1, PC2, colour = pop)) +
  geom_point(size = 3) +
  labs(x = sprintf("PCA1 (%.1f%%)", var_exp[1]),
       y = sprintf("PCA2 (%.1f%%)", var_exp[2])) +
  theme_classic()

# Create interactive 3D PCA plot with plotly
pca_3d <- plot_ly(scores, 
                  x = ~PC1, 
                  y = ~PC2, 
                  z = ~PC3,
                  color = ~pop,
                  colors = my.colors1,
                  type = "scatter3d",
                  mode = "markers",
                  marker = list(size = 5, opacity = 0.8),
                  text = ~paste("Population:", pop),
                  hoverinfo = "text") %>%
  layout(
    title = "Interactive 3D PCA of Population Structure",
    scene = list(
      xaxis = list(title = paste0("PC1 (", round(100*pca1$eig[1]/sum(pca1$eig),1), "%)")),
      yaxis = list(title = paste0("PC2 (", round(100*pca1$eig[2]/sum(pca1$eig),1), "%)")),
      zaxis = list(title = paste0("PC3 (", round(100*pca1$eig[3]/sum(pca1$eig),1), "%)"))
    )
  )

pca_3d

# Same results as that of the PCoA.


# Save as HTML widget (interactive)
htmlwidgets::saveWidget(pca_3d, "pca_3d_interactive.html")
```




# DAPC ANALYSES
## run dapc based of our knowledge from the PCA above
```{r}
paxcan.dapc <- dapc(paxcan_gl, n.pca = 3, n.da = 2)
```



## plot the dapc
```{r}
scatter(paxcan.dapc, 
        col = my.colors1, 
        cex = 2, 
        legend = TRUE, 
        clabel = F, 
        posi.leg = "topleft", scree.pca = TRUE,
        posi.pca = "bottomleft", cleg = 0.75)


# compoplot
dapc.results <- as.data.frame(paxcan.dapc$posterior)
dapc.results$pop <- pop(paxcan_gl)
dapc.results$indNames <- rownames(dapc.results)

# make the dataframe longer for easy visualization
dapc.results <- pivot_longer(dapc.results, -c(pop, indNames))

# rename the columns
colnames(dapc.results) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")

# create the compoplot
p <- ggplot(dapc.results, 
            aes(x = Sample, y = Posterior_membership_probability, fill = Assigned_Pop)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = my.colors1) +
  facet_grid(~Original_Pop, scales = "free") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8)
  )
p

# i am tempted to agree with scott schuette on the where ancestral populations lie across the range, this feature seen is similar to the theory on populations on the extremes of the species range being very dissimilar from the others.
```







