---
title: "paxcan_basic_stats_analyses"
author: "Isaac Buabeng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## load the necessary libraires
```{r}
library(hierfstat)
library(gaston)
library(vcfR)
library(reshape2)
library(pegas)
library(gt)
library(cowplot)
library(ggplot2)
library(tidyverse)
```



# load in the data
```{r}
vcf <- read.vcfR("14pops_biallelic_snps_miss0.4_thin10.newID.mac1.maf0.01.ldkb1r0.8.vcf")


# load in population data
# cords for each individual
all_cords <- read.csv("coords_all.csv")

# cords for each population
cords <- read.csv("coords.csv")
```


# calculate basic fstats
## convert the vcf file into a genind object for hierfstsat
```{r vcf2genind}
paxcan_genind <- vcfR2genind(vcf)
```



## convert the vcf object to a genlight object for dartR
```{r}
paxcan_gl <- vcfR2genlight(vcf)

# Ensure all required metadata and flags exist
paxcan_gl <- gl.compliance.check(paxcan_gl)   # adds missing flag skeletons

# Remove (or confirm absence of) monomorphic loci and set the flag
paxcan_gl <- gl.filter.monomorphs(paxcan_gl, verbose = 3)

# assign population names
pop(paxcan_gl) <- as.factor(all_cords$NAME_OF_POPULATION)

popNames(paxcan_gl)
```



## assign population information
```{r}
pop(paxcan_genind) <- as.factor(all_cords$NAME_OF_POPULATION)

popNames(paxcan_genind)
# perfect!
```



## convert the genind object into hierfstat format
```{r}
paxcan_hfstat <- genind2hierfstat(paxcan_genind, pop=NULL)
```



## calculate the basic fstats with hierfstat
```{r}
basicstat <- basic.stats(paxcan_hfstat, diploid=TRUE, digits=3)

# inspect the results
basicstat$overall

# save the results
save(basicstat, file="basicstat.RData")
```


## confirm with dartR
```{r}
gl.basic.stats(paxcan_gl)
# very similar to what we saw in hierfstat
```



## save the results into a dataframe
```{r}
overall_basic_stats <- as.data.frame(basicstat$overall)

# examine the table
overall_basic_stats
```



## Get bootstrap confidence values for Fis
```{r}
boot <- boot.ppfis(paxcan_hfstat, nboot=1000)
boot5 <-boot.ppfis(paxcan_hfstat, nboot=1000, quant=0.5)

boot
boot5
```



## Make a table for our bootstrapped observed and expected heterozygosity values
```{r}
# extract population names for constructing the dataframe
pop_order <- cords$NAME_OF_POPULATION

# use the names of the populations to rearrange the heterozygosity results
colnames(basicstat$Ho) <- pop_order
```



## construct the table for fis values
```{r}
Ho <- colMeans(basicstat$Ho,na.rm=T)
He <- colMeans(basicstat$Hs,na.rm=T)
Fis <- boot5$fis.ci$ll
new_popgen_stats <- cbind(Ho,He,Fis,boot$fis.ci)

new_popgen_stats
```



## save the resulst
```{r}
popgen_stats_noMD.csv <-write.table(new_popgen_stats, sep=",", dec=".", row.names=TRUE, col.names=TRUE)

save(new_popgen_stats, file="new_popgen_stats")
```



## calculate fst 
```{r}
paxcan.fst.mat <- genet.dist(paxcan_hfstat, method = "WC84")

gindF.fst.mat.tri <- as.matrix(paxcan.fst.mat)

colnames(gindF.fst.mat.tri) <- pop_order

rownames(gindF.fst.mat.tri) <- pop_order

write.table(gindF.fst.mat.tri, file="paxcanFstWC84.pwfst", row.names=T,col.names=T)
```



## construct a table for the results
```{r}
melted <- melt(gindF.fst.mat.tri, na.rm = TRUE)
```


## look at results
```{r}
summary(melted$value)
```



## pairwise fst plot
```{r}
combined <- ggplot(data = melted, aes(Var2, Var1, fill = value))+ 
  geom_tile(color = "white")+ 
  scale_fill_gradient(low = "white", high = "red", name="FST")  +
  ggtitle(expression(atop("Pairwise FST, WC (1984), Combined", atop(italic("N = 118"), ""))))+
  labs( x = "Sampling Site", y = "Sampling Site") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1),axis.text.y = element_text(size = 13)) + 
  theme_classic()+
  coord_fixed()
combined
```



